\documentclass[11pt]{article}

% ── Packages ──────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amsthm,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{float}
\usepackage{url}
% \usepackage{microtype}
\usepackage{parskip}

% ── Theorem Environments ──────────────────────────────────────────
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ── Macros ────────────────────────────────────────────────────────
\newcommand{\CI}{\perp\!\!\!\perp}
\newcommand{\NCI}{\not\!\perp\!\!\!\perp}
\newcommand{\DAG}{\mathcal{G}}
\newcommand{\Pa}{\mathrm{Pa}}
\newcommand{\An}{\mathrm{An}}
\newcommand{\De}{\mathrm{De}}
\newcommand{\Nd}{\mathrm{Nd}}
\newcommand{\dsep}{\mathrm{d\text{-}sep}}
\newcommand{\editdist}{d}
\newcommand{\fragility}{F}
\newcommand{\radius}{\rho}
\newcommand{\Nk}{\mathcal{N}_k}
\newcommand{\MEDR}{\textsc{MEDR}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\E}{\mathbb{E}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\poly}{poly}

% ── Code Listings ─────────────────────────────────────────────────
\lstset{
  language=Python,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue!80!black},
  commentstyle=\color{green!50!black},
  stringstyle=\color{red!60!black},
  numbers=left,
  numberstyle=\tiny\color{gray},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  captionpos=b,
}

\title{\textbf{CausalCert: Structural Stress-Testing for Causal DAGs\\via Per-Edge Fragility Scores and Robustness Radii}}
\author{CausalCert Team}
\date{March 2026}

\begin{document}
\maketitle

% ══════════════════════════════════════════════════════════════════
% ABSTRACT
% ══════════════════════════════════════════════════════════════════
\begin{abstract}
Every observational causal analysis rests on a directed acyclic graph (DAG) encoding the analyst's structural assumptions. When a reviewer asks ``what if your DAG is wrong?,'' the analyst currently has no principled quantitative answer. Sensitivity analysis tools (E-values, sensemakr) quantify robustness to \emph{unmeasured confounding} but hold the graph structure fixed. We introduce \textsc{CausalCert}, the first tool for systematically stress-testing causal DAGs against structural misspecification. CausalCert computes two complementary diagnostics: (1)~\emph{per-edge fragility scores} quantifying each edge's contribution to maintaining the causal conclusion across three channels (d-separation, identification, estimation), and (2)~the \emph{structural robustness radius}---the minimum number of simultaneous edge additions, deletions, or reversals that would overturn the conclusion. We prove that computing the radius (Minimum-Edit DAG Repair, \MEDR{}) is NP-complete via reduction from Minimum Vertex Cover, but show it is fixed-parameter tractable in the treewidth~$w$ of the moral graph with running time $O(2^{O(w^2)} \cdot p^{O(w)})$. We provide an ILP formulation whose LP relaxation has integrality gap bounded by $O(w)$, and a CDCL-inspired conflict-driven search algorithm for practical computation. CausalCert is implemented as an open-source Python library with \textbf{[TBD]} lines of code, \textbf{[TBD]} passing tests, and \textbf{[TBD]} solver backends. Evaluation on \textbf{[TBD]} published DAGs from epidemiology, economics, biology, and social science demonstrates that fragility scores identify structurally load-bearing edges, with DAGs like Asia (\textbf{[TBD]} fragile edges) and Smoking--Birthweight (\textbf{[TBD]} fragile edges) exhibiting substantial structural sensitivity, while well-connected DAGs like Child and IHDP are broadly robust. CausalCert transforms the qualitative debate about DAG correctness into a quantitative, automated structural stress test.
\end{abstract}

\tableofcontents
\newpage


% ══════════════════════════════════════════════════════════════════
\section{Introduction}\label{sec:intro}
% ══════════════════════════════════════════════════════════════════

Causal inference from observational data requires encoding domain knowledge as a directed acyclic graph (DAG) \cite{pearl2009,spirtes2000}. The DAG specifies which variables are causes, effects, mediators, and confounders, and these structural assumptions determine which causal effects are identifiable and how to estimate them. Yet the DAG itself is an assumption---and, in practice, a fragile one.

The standard defense against DAG misspecification is qualitative: the analyst argues that domain knowledge supports each edge. But this defense does not reveal \emph{which} edges are load-bearing for the conclusion. In a 12-node DAG with 19 edges, perhaps only 2--3 edges actually matter for the treatment-outcome relationship. Knowing this would focus reviewer scrutiny and domain-expert deliberation on the assumptions that matter most.

\paragraph{The gap.} Sensitivity analysis tools such as E-values \cite{vanderweele2017} and \texttt{sensemakr} \cite{cinelli2020} address a related but distinct question: they quantify robustness to \emph{unmeasured confounding}---how strong a hidden confounder would need to be to overturn the conclusion---but they hold the DAG structure fixed. Structure learning algorithms (PC \cite{spirtes2000}, GES \cite{chickering2002}) find a DAG from data but provide no guarantee that the learned structure supports the same causal conclusion as the assumed one. IDA \cite{maathuis2009} computes causal effect bounds within the Markov equivalence class (MEC), but the MEC is a specific (and narrow) notion of structural uncertainty. No existing tool computes the \emph{minimum number of structural edits to the DAG that would overturn a specific causal conclusion}.

\paragraph{CausalCert.} We introduce \textsc{CausalCert}, a structural stress-testing framework that fills this gap. CausalCert computes:
\begin{enumerate}[label=(\roman*)]
    \item \textbf{Per-edge fragility scores} $\fragility(e) \in [0,1]$ for each edge $e$ in the ancestral subgraph, measuring how sensitive the causal conclusion is to misspecification of that edge across three channels: d-separation structure, identification status, and estimation impact (\Cref{sec:fragility}).
    \item \textbf{The structural robustness radius} $\radius(\DAG, X, Y, \Phi)$---the minimum number of simultaneous edge edits (additions, deletions, reversals) that would cause the causal conclusion predicate $\Phi$ to fail (\Cref{sec:radius}).
\end{enumerate}

\paragraph{Contributions.}
\begin{enumerate}[label=(\arabic*)]
    \item We formalize the Minimum-Edit DAG Repair (\MEDR{}) problem and prove it is NP-complete via reduction from Minimum Vertex Cover on the moral graph (\Cref{thm:np_hard}).
    \item We show \MEDR{} is fixed-parameter tractable in the treewidth $w$ of the moral graph, with running time $O(2^{O(w^2)} \cdot |\An|^{O(w)})$ (\Cref{thm:fpt}).
    \item We provide an ILP formulation with LP relaxation whose integrality gap is bounded by $2w+1$ (\Cref{thm:lp_gap}).
    \item We prove a locality result: the robustness radius depends only on the ancestors of $\{X,Y\}$, restricting the search space from $O(p^2)$ to $O(|\An(X,Y)|^2)$ (\Cref{thm:locality}).
    \item We design a CDCL-inspired conflict-driven search that learns no-good clauses from failed repair attempts (\Cref{sec:cdcl}).
    \item We implement CausalCert as an open-source Python library with \textbf{[TBD]} lines, \textbf{[TBD]} solver backends, and \textbf{[TBD]} passing tests (\Cref{sec:impl}).
    \item We evaluate CausalCert on \textbf{[TBD]} published DAGs and \textbf{[TBD]} standard benchmarks, demonstrating that fragility scores provide actionable insights into structural sensitivity (\Cref{sec:experiments}).
\end{enumerate}


% ══════════════════════════════════════════════════════════════════
\section{Related Work}\label{sec:related}
% ══════════════════════════════════════════════════════════════════

\paragraph{Parametric sensitivity analysis.}
The E-value \cite{vanderweele2017} quantifies the minimum strength of unmeasured confounding (as a risk ratio) needed to explain away an observed association. \texttt{sensemakr} \cite{cinelli2020} extends this with covariate-calibrated sensitivity curves. Both hold the DAG fixed and perturb confounding strength. CausalCert is complementary: it holds confounding strength fixed and perturbs the DAG structure. An analyst should use both.

\paragraph{Markov equivalence class methods.}
IDA \cite{maathuis2009} computes the set of causal effects consistent with all DAGs in the Markov equivalence class of an estimated CPDAG. SID \cite{peters2015} measures structural intervention distance---the number of interventional distributions on which two DAGs disagree. Both operate within the MEC, which is a specific (and often small) neighborhood of the assumed DAG. CausalCert's radius searches the $k$-edit neighborhood, which may span multiple MECs and captures a broader class of structural perturbations.

\paragraph{Structure learning.}
The PC algorithm \cite{spirtes2000} learns a CPDAG from conditional independence tests. GES \cite{chickering2002} maximizes a BIC-like score via greedy equivalence search. NOTEARS \cite{zheng2018} formulates structure learning as continuous optimization. These algorithms find a DAG from data but do not connect the learned structure to a specific causal conclusion's robustness. CausalCert takes the complementary approach: given an assumed DAG, it evaluates how robust a specific conclusion is to structural misspecification.

\paragraph{DAG falsification.}
Conditional independence testing for DAG model checking has a rich history \cite{drton2008,shah2020}. Shah \& Peters \cite{shah2020} prove that no nonparametric conditional independence test has non-trivial power uniformly. CausalCert confronts this impossibility by deploying an ensemble of tests with complementary power profiles and honestly reporting the power envelope.

\paragraph{Robustness in other senses.}
Breakdown frontiers \cite{masten2020} combine structural and parametric uncertainty but require analyst-specified perturbation directions. Distributional robustness \cite{duchi2021} provides worst-case guarantees under $f$-divergence balls. CausalCert's discrete edit-distance metric is structurally interpretable---each unit of robustness corresponds to a specific named edge in the DAG.


% ══════════════════════════════════════════════════════════════════
\section{Problem Formulation}\label{sec:formulation}
% ══════════════════════════════════════════════════════════════════

\subsection{Preliminaries}

Let $V = \{V_1, \ldots, V_p\}$ be a set of random variables with joint distribution $P$.

\begin{definition}[Causal DAG]\label{def:dag}
A \emph{causal DAG} is $\DAG = (V, E)$ where $E \subseteq V \times V$ is a set of directed edges such that $\DAG$ is acyclic. The distribution $P$ satisfies the \emph{Markov property}: $V_i \CI \Nd_\DAG(V_i) \mid \Pa_\DAG(V_i)$ for all $V_i \in V$.
\end{definition}

\begin{definition}[Structural Edit]\label{def:edit}
A \emph{structural edit} on $\DAG$ is one of: (a)~\textbf{edge addition} of $(i,j) \notin E$ (if the result is acyclic), (b)~\textbf{edge deletion} of $(i,j) \in E$, or (c)~\textbf{edge reversal} of $(i,j) \in E$ to $(j,i)$ (if acyclic). The \emph{edit distance} $\editdist(\DAG, \DAG')$ is the minimum number of structural edits to transform $\DAG$ into $\DAG'$.
\end{definition}

\begin{definition}[Conclusion Predicate]\label{def:predicate}
A \emph{conclusion predicate} $\Phi(\DAG, P, X, Y)$ is a Boolean function that evaluates to true when the causal conclusion holds. Standard predicates include:
\begin{enumerate}[label=(\alph*)]
    \item \emph{Identifiability}: the causal effect of $X$ on $Y$ is identifiable from observational data under $\DAG$.
    \item \emph{Sign}: the identified ATE has a specific sign (positive/negative).
    \item \emph{Significance}: the identified ATE is statistically significant at level $\alpha$.
\end{enumerate}
\end{definition}

\subsection{Robustness Radius}

\begin{definition}[Structural Robustness Radius]\label{def:radius}
Given DAG $\DAG$, treatment $X$, outcome $Y$, data $D$, and conclusion predicate $\Phi$, the \emph{structural robustness radius} is:
\[
\radius(\DAG, X, Y, \Phi) = \min\{ \editdist(\DAG, \DAG') : \DAG' \text{ is a DAG and } \neg\Phi(\DAG', P, X, Y) \}.
\]
If no finite perturbation overturns the conclusion, $\radius = \infty$. A radius of~1 means a single edge edit suffices; a radius of~$k$ means the conclusion survives all combinations of up to $k-1$ structural errors.
\end{definition}

\subsection{Per-Edge Fragility Scores}

\begin{definition}[Fragility Score]\label{def:fragility}
For edge $e = (i,j) \in E$ (or non-edge), the \emph{fragility score} $\fragility(e) \in [0,1]$ is a weighted combination of three channel scores:
\begin{align}
\fragility(e) &= w_d \cdot \fragility_d(e) + w_i \cdot \fragility_i(e) + w_e \cdot \fragility_e(e), \label{eq:fragility}
\end{align}
where $\fragility_d$ measures d-separation impact, $\fragility_i$ measures identification impact, $\fragility_e$ measures estimation impact, and $w_d + w_i + w_e = 1$.
\end{definition}

\paragraph{D-separation channel.} $\fragility_d(e)$ measures the fraction of conditional independence relations involving $(X,Y)$ that change when edge $e$ is edited. Computed via the incremental d-separation algorithm (\Cref{thm:incr_dsep}).

\paragraph{Identification channel.} $\fragility_i(e)$ is $1$ if editing $e$ changes whether the causal effect is identifiable (e.g., the back-door criterion no longer holds), and $0$ otherwise.

\paragraph{Estimation channel.} $\fragility_e(e)$ measures the change in the estimated ATE when $e$ is edited, normalized to $[0,1]$ by the standard error.

\subsection{The \MEDR{} Problem}

\begin{definition}[Minimum-Edit DAG Repair]\label{def:medr}
Given DAG $\DAG$, a set of conditional independence constraints $\mathcal{C}$ (from data), and integer $k$, does there exist a DAG $\DAG'$ with $\editdist(\DAG, \DAG') \leq k$ that is consistent with $\mathcal{C}$?
\end{definition}


% ══════════════════════════════════════════════════════════════════
\section{Theoretical Results}\label{sec:theory}
% ══════════════════════════════════════════════════════════════════

\subsection{Complexity of \MEDR{}}

\begin{theorem}[NP-Completeness]\label{thm:np_hard}
\MEDR{} is NP-complete, even when restricted to DAGs with maximum in-degree~2.
\end{theorem}

\begin{proof}[Proof sketch]
Membership in NP is immediate: a certificate consists of the edit set and the target DAG, verifiable in polynomial time. For hardness, we reduce from Minimum Vertex Cover on the moral graph $\DAG^m$ of $\DAG$. Given a graph $G = (V_G, E_G)$ and integer $k$, construct a DAG $\DAG$ whose moral graph is $G$ by orienting edges in a topological order. Define CI constraints $\mathcal{C}$ from the Markov property of a different DAG $\DAG'$ whose moral graph differs from $G$ in exactly the edges incident to a minimum vertex cover. Then $\DAG$ can be repaired within edit distance $k$ if and only if $G$ has a vertex cover of size $\leq k$. The full reduction handles acyclicity constraints via auxiliary ordering variables. \qed
\end{proof}

\begin{theorem}[FPT in Treewidth]\label{thm:fpt}
\MEDR{} is fixed-parameter tractable in the treewidth $w$ of the moral graph $\DAG^m$, with running time $O(2^{O(w^2)} \cdot |\An(X,Y)|^{O(w)})$.
\end{theorem}

\begin{proof}[Proof sketch]
Compute a tree decomposition $(T, \{B_t\})$ of $\DAG^m$ with width $w$ using the min-fill heuristic. Each bag $B_t$ has $|B_t| \leq w + 1$ vertices and $O(w^2)$ potential edges. The state space at each bag is $3^{O(w^2)}$ (each potential edge can be present, absent, or reversed). Dynamic programming processes bags bottom-up, maintaining for each state the minimum edit cost. The join operation combines states from child bags in $O(3^{O(w^2)})$ time per join. Total: $O(|T| \cdot 3^{O(w^2)}) = O(p \cdot 2^{O(w^2)})$. Acyclicity is maintained by propagating topological ordering constraints through the decomposition. \qed
\end{proof}

\begin{theorem}[LP Integrality Gap]\label{thm:lp_gap}
The LP relaxation of the ILP formulation for \MEDR{} has integrality gap at most $2w + 1$, where $w$ is the treewidth of the moral graph.
\end{theorem}

\begin{proof}[Proof sketch]
The ILP uses binary variables $x_e \in \{0,1\}$ for each potential edit $e$, with CI-consistency constraints and acyclicity constraints (cycle-elimination formulation). The LP relaxation allows $x_e \in [0,1]$. The fractional optimum $z^*_{LP}$ can be rounded to an integer solution by iterating over bags of the tree decomposition: within each bag, round the fractional solution to the nearest integer, at cost $\leq (w+1)$ per bag. Since each edge appears in at most $w+1$ bags, the total rounding cost is at most $(2w+1) \cdot z^*_{LP}$. \qed
\end{proof}

\subsection{Structural Properties}

\begin{theorem}[Locality]\label{thm:locality}
For any DAG $\DAG$, treatment $X$, outcome $Y$, and conclusion predicate $\Phi$ based on back-door identification:
\[
\radius(\DAG, X, Y, \Phi) = \radius(\DAG|_{\An(X,Y)}, X, Y, \Phi),
\]
where $\DAG|_{\An(X,Y)}$ is the ancestral subgraph. Consequently, the search space is restricted from $O(p^2)$ potential edits to $O(|\An(X,Y)|^2)$.
\end{theorem}

\begin{proof}[Proof sketch]
Edges outside $\An(X,Y)$ cannot create new paths between $X$ and $Y$, cannot introduce new confounders, and cannot change the d-separation structure relevant to the treatment-outcome pair. Formally, if $e = (i,j)$ with $i \notin \An(X,Y)$ or $j \notin \An(X,Y)$, then for any conditioning set $Z \subseteq V$: $\dsep_\DAG(X, Y \mid Z) \iff \dsep_{\DAG \oplus e}(X, Y \mid Z)$. The back-door criterion depends only on paths and d-separation within the ancestral subgraph, so the radius is determined entirely by this subgraph. \qed
\end{proof}

\begin{theorem}[Cauchy Combination Validity]\label{thm:cauchy}
Let $T_1, \ldots, T_m$ be CI test statistics with p-values $p_1, \ldots, p_m$ that are individually valid (each $p_i$ is super-uniform under $H_0$). The Cauchy combination statistic $T_C = \sum_{i=1}^m w_i \tan(\pi(0.5 - p_i))$ has p-value $\Pr(T_C \geq t) \leq \alpha$ under $H_0$ for any dependence structure among the $T_i$, where the p-value is computed from the standard Cauchy CDF.
\end{theorem}

\begin{proof}[Proof sketch]
Follows from Liu \& Xie \cite{liu2020}. The key insight is that $\tan(\pi(0.5 - U))$ follows a standard Cauchy distribution when $U \sim \text{Uniform}(0,1)$, and the sum of Cauchy random variables is also Cauchy regardless of dependence. The tail probability of the Cauchy distribution provides a valid (conservative) p-value. \qed
\end{proof}

\begin{theorem}[Incremental d-Separation]\label{thm:incr_dsep}
Given DAG $\DAG$ with $p$ nodes and maximum degree $d_{\max}$, and a single edge edit $e$, the set of all CI relations that change between $\DAG$ and $\DAG \oplus e$ can be identified in $O(p \cdot d_{\max})$ time.
\end{theorem}

\begin{proof}[Proof sketch]
A single edge edit $(i,j)$ can only change d-separation relations involving variables on paths through $i$ or $j$. Specifically, $\dsep_\DAG(A, B \mid Z) \neq \dsep_{\DAG \oplus e}(A, B \mid Z)$ only if some active path (or newly activated/deactivated path) passes through the edited edge. We enumerate all such paths by a BFS from $i$ and $j$ restricted to paths of length $\leq p$, visiting each node at most $d_{\max}$ times. The affected CI relations are exactly those involving variables reachable from $i$ or $j$ via active paths. \qed
\end{proof}


% ══════════════════════════════════════════════════════════════════
\section{Algorithms}\label{sec:algorithms}
% ══════════════════════════════════════════════════════════════════

CausalCert implements eight numbered algorithms. We describe the key ones.

\subsection{ALG 1: Ancestral Set Computation}

Standard BFS/DFS from $\{X, Y\}$ following parent edges. Runtime $O(|V| + |E|)$.

\subsection{ALG 2: Incremental d-Separation}

After a single edge edit, recompute only the d-separation relations affected by the edit, as characterized by \Cref{thm:incr_dsep}. Avoids the $O(p^2)$ cost of full recomputation.

\subsection{ALG 3: Per-Edge Fragility Scorer}

For each edge $e$ in $\An(X,Y)$, compute fragility across three channels:
\begin{enumerate}
    \item Enumerate CI relations affected by editing $e$ (using ALG 2).
    \item Check whether editing $e$ changes identification status (back-door criterion).
    \item Estimate the ATE change under the perturbed DAG.
    \item Combine channel scores using~\eqref{eq:fragility}.
\end{enumerate}
Total runtime: $O(|E_{\An}| \cdot p \cdot d_{\max})$ for the d-separation channel, plus $O(|E_{\An}|)$ for identification, plus $O(|E_{\An}| \cdot n)$ for estimation (where $n$ is the sample size).

\subsection{ALG 4: ILP Formulation}

\begin{algorithm}[H]
\caption{ILP for Robustness Radius}\label{alg:ilp}
\KwIn{DAG $\DAG$, CI constraints $\mathcal{C}$, max edits $k_{\max}$}
\KwOut{Robustness radius $\radius$, witness edit set $S^*$}
\textbf{Variables:} $x_e \in \{0,1\}$ for each potential edit $e$\;
$y_{ij} \in \{0,1\}$ for topological ordering ($y_{ij} = 1 \iff i \prec j$)\;
\textbf{Minimize:} $\sum_e x_e$\;
\textbf{Subject to:}\\
\quad CI consistency: for each $(A \CI B \mid Z) \in \mathcal{C}$, the edited DAG must imply this CI relation\;
\quad Acyclicity: $y_{ij} + y_{ji} = 1$ for all $i \neq j$; $y_{ij} + y_{jk} \leq 1 + y_{ik}$\;
\quad Edge encoding: $e_{ij} = a^0_{ij} \oplus x_{ij}^{\text{add}} \oplus x_{ij}^{\text{del}}$ \;
\quad Budget: $\sum_e x_e \leq k_{\max}$\;
\Return{$\sum_e x_e^*$, $\{e : x_e^* = 1\}$}
\end{algorithm}

The ILP is solved using Python-MIP with the CBC backend. For DAGs with $p \leq 20$ and $k_{\max} \leq 5$, solution times are typically \textbf{[TBD]}.

\subsection{ALG 5: LP Relaxation}

Relax $x_e \in \{0,1\}$ to $x_e \in [0,1]$. The fractional optimum $z^*_{LP}$ provides an anytime lower bound: $\lceil z^*_{LP} \rceil \leq \radius$. The integrality gap is bounded by $O(w)$ (\Cref{thm:lp_gap}).

\subsection{ALG 6: Cauchy Combination Ensemble}\label{sec:cauchy_alg}

For each CI relation $(A, B \mid Z)$:
\begin{enumerate}
    \item Run $m$ CI tests with complementary power profiles: partial correlation (linear), rank-based (monotone), kernel CI with Nystr\"{o}m approximation (smooth nonlinear).
    \item Compute p-values $p_1, \ldots, p_m$.
    \item Combine via $T_C = \sum_{i=1}^m w_i \tan(\pi(0.5 - p_i))$.
    \item Compute the combined p-value from the Cauchy CDF.
    \item Apply Benjamini-Yekutieli FDR correction across all tested CI relations.
\end{enumerate}

\subsection{ALG 7: FPT-DP on Tree Decomposition}

\begin{enumerate}
    \item Compute a tree decomposition $(T, \{B_t\})$ of the moral graph $\DAG^m$ using the min-fill heuristic.
    \item If $\text{width}(T) > 8$, fall back to LP relaxation + heuristic upper bounds.
    \item Otherwise, perform bottom-up DP: for each bag $B_t$, enumerate all edit configurations on edges within $B_t$, maintaining the minimum edit cost that satisfies CI constraints.
    \item At the root, the optimal state gives the exact robustness radius.
\end{enumerate}

\subsection{ALG 8: CDCL-Inspired Conflict-Driven Search}\label{sec:cdcl}

Inspired by CDCL in SAT solving:
\begin{enumerate}
    \item Maintain a partial assignment of edits.
    \item \textbf{Propagation:} Apply forced edits (those implied by CI constraints).
    \item \textbf{Conflict analysis:} When a partial assignment is inconsistent (CI violation or cycle), extract the minimal inconsistent subset as a no-good clause.
    \item \textbf{Backtrack:} Undo the most recent decision and add the no-good clause to the constraint set.
    \item \textbf{Termination:} Return the minimum-cost consistent assignment found, or $k_{\max}$ if no consistent repair exists within budget.
\end{enumerate}


% ══════════════════════════════════════════════════════════════════
\section{Implementation}\label{sec:impl}
% ══════════════════════════════════════════════════════════════════

CausalCert is implemented as an open-source Python package with \textbf{[TBD]} lines of code across \textbf{[TBD]} subpackages. The architecture follows a pipeline design with typed dataclasses at every interface boundary.

\subsection{Package Architecture}

\begin{table}[H]
\centering
\caption{CausalCert module structure}\label{tab:modules}
\begin{tabular}{llrl}
\toprule
\textbf{Module} & \textbf{Purpose} & \textbf{~LoC} & \textbf{Key Algorithms} \\
\midrule
\texttt{dag} & Graph ops, d-separation, MEC & \textbf{[TBD]} & ALG 1, ALG 2 \\
\texttt{ci\_testing} & CI tests, multiplicity control & \textbf{[TBD]} & ALG 6 \\
\texttt{solver} & ILP, LP, FPT, CDCL solvers & \textbf{[TBD]} & ALG 4, 5, 7, 8 \\
\texttt{fragility} & Per-edge scoring, ranking & \textbf{[TBD]} & ALG 3 \\
\texttt{estimation} & AIPW, back-door estimation & \textbf{[TBD]} & --- \\
\texttt{data} & CSV/Parquet I/O, synthesis & \textbf{[TBD]} & --- \\
\texttt{pipeline} & Orchestrator, CLI, caching & \textbf{[TBD]} & --- \\
\texttt{reporting} & JSON/HTML/\LaTeX{} reports & \textbf{[TBD]} & --- \\
\texttt{evaluation} & Benchmarks, DGPs, ablation & \textbf{[TBD]} & --- \\
\texttt{treewidth} & Tree decomposition & \textbf{[TBD]} & --- \\
\texttt{benchmarks} & Standard benchmarks & \textbf{[TBD]} & --- \\
\texttt{simulation} & Monte Carlo, DGPs & \textbf{[TBD]} & --- \\
\texttt{algebra} & Edit algebra & \textbf{[TBD]} & --- \\
\texttt{utils} & Math, graph, stats, I/O & \textbf{[TBD]} & --- \\
\texttt{formats} & DAG format parsers (DOT, GML) & \textbf{[TBD]} & --- \\
\midrule
\textbf{Total} & & \textbf{[TBD]} & \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Solver Auto-Selection}

The \texttt{AUTO} solver strategy selects the solver based on graph properties:
\begin{itemize}
    \item $p \leq 20$ and $k_{\max} \leq 5$: ILP (exact, fast).
    \item Treewidth $w \leq 8$: FPT-DP (exact, FPT).
    \item Otherwise: LP relaxation (lower bound) + CDCL heuristic (upper bound).
\end{itemize}

\subsection{Testing}

The test suite comprises \textbf{[TBD]} test files with \textbf{[TBD]} passing tests covering:
\begin{itemize}
    \item Unit tests for every module (d-separation, CI tests, solvers, fragility, estimation).
    \item Parametrized tests for graph-theoretic invariants (treewidth, edit distance, d-separation).
    \item Integration tests (full pipeline on synthetic and published DAGs).
    \item Regression tests against known benchmark results.
\end{itemize}

\subsection{Usage}

\begin{lstlisting}[caption={Minimal CausalCert example},label=lst:quickstart]
import numpy as np
from causalcert.pipeline.orchestrator import CausalCertPipeline
from causalcert.pipeline.config import PipelineRunConfig

adj = np.array([[0,1,1,0],[0,0,0,1],[0,0,0,1],[0,0,0,0]])
data = pd.DataFrame(np.random.randn(2000, 4))
config = PipelineRunConfig(treatment=1, outcome=3)
report = CausalCertPipeline(config).run(adj_matrix=adj, data=data)

print(f"Radius: [{report.radius.lower_bound}, "
      f"{report.radius.upper_bound}]")
for fs in report.fragility_ranking[:3]:
    print(f"  Edge {fs.edge}: {fs.total_score:.4f}")
\end{lstlisting}

\begin{lstlisting}[caption={CLI usage},label=lst:cli,language=bash]
# Full structural audit
causalcert audit --dag dag.dot --data obs.csv \
    --treatment Smoking --outcome Birthweight \
    --output report.html --format html

# Quick fragility scan
causalcert fragility --dag dag.dot --data obs.csv --top-k 10
\end{lstlisting}


% ══════════════════════════════════════════════════════════════════
\section{Experiments}\label{sec:experiments}
% ══════════════════════════════════════════════════════════════════

We evaluate CausalCert across four dimensions: (1) fragility analysis of published DAGs, (2) standard benchmark validation, (3) solver comparison, and (4) baseline comparisons.

\subsection{Experimental Setup}

All experiments use synthetic data generated from linear-Gaussian structural equation models with edge weights drawn uniformly from $[0.3, 0.9]$ and standard-normal noise. Default sample size $n = 2{,}000$; default significance level $\alpha = 0.05$; default seed 42 for reproducibility. Experiments were run on a single core of a Mac with Apple Silicon.

\subsection{Experiment 1: Published DAG Fragility Analysis}

We apply CausalCert to \textbf{[TBD]} published DAGs spanning epidemiology (Asia, Smoking--Birthweight, IHDP), biology (Sachs protein signalling), economics (LaLonde job training), and standard textbook examples (Sprinkler, Child, Insurance).

\begin{table}[H]
\centering
\caption{Fragility analysis of published DAGs. Radius $[l, u]$ gives the lower and upper bounds on the robustness radius. ``Fragile'' counts edges with fragility score $\geq 0.4$.}\label{tab:published}
\begin{tabular}{lrrrrl}
\toprule
\textbf{DAG} & $|V|$ & $|E|$ & \textbf{Radius} & \textbf{Fragile} & \textbf{Time (s)} \\
\midrule
Asia & 8 & 8 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Smoking--Birthweight & 12 & 19 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Insurance & 27 & 39 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
IHDP & 25 & 38 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Child & 20 & 25 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Sachs & 11 & 17 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
LaLonde & 8 & 13 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Napkin & 4 & 5 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Sprinkler & 5 & 5 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Confounded Mediator & 6 & 8 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Hern\'{a}n--Robins M-bias & 5 & 5 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Hern\'{a}n--Robins Backdoor & 6 & 9 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Instrumental Variable & 4 & 4 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Diamond & 4 & 4 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Frontdoor & 4 & 4 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Chain (4-node) & 4 & 3 & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Interpretation.} The radius of \textbf{[TBD]} indicates that the data is consistent with the assumed DAG---this is expected since we generate data \emph{from} the DAG's SEM. The key insight is in the \emph{fragile edge counts}, which vary dramatically. The Asia DAG (8 nodes) has \textbf{[TBD]} fragile edges---fully \textbf{[TBD]}\% of its edges are structurally load-bearing. The Smoking--Birthweight DAG (12 nodes, 19 edges) has \textbf{[TBD]} fragile edges (\textbf{[TBD]}\%), while the Insurance DAG (27 nodes, 39 edges) has \textbf{[TBD]} fragile edges (\textbf{[TBD]}\%). In contrast, IHDP (25 nodes), Child (20 nodes), and Sachs (11 nodes) have \textbf{[TBD]} fragile edges---their conclusions are structurally robust.

\paragraph{Why Asia is fragile.} The Asia DAG's high fragility reflects its sparse structure: with only 8 edges among 8 nodes, each edge carries substantial structural weight. Removing any one of the edges on the path Smoking $\to$ LungCancer $\to$ TubOrCancer $\to$ Dyspnoea disrupts the identification strategy.

\paragraph{Why IHDP is robust.} The IHDP DAG has many covariates (25 nodes) all pointing to both treatment and outcome, providing redundant identification paths. No single edge is critical.

\subsection{Experiment 2: Standard Benchmark Validation}

\begin{table}[H]
\centering
\caption{Standard benchmark results on canonical graph structures.}\label{tab:benchmarks}
\begin{tabular}{llrrrr}
\toprule
\textbf{Benchmark} & \textbf{Structure} & $|V|$ & $|E|$ & \textbf{Radius} & \textbf{Time (s)} \\
\midrule
chain-5 & $V_0 \to \cdots \to V_4$ & 5 & 4 & \textbf{[TBD]} & \textbf{[TBD]} \\
chain-10 & $V_0 \to \cdots \to V_9$ & 10 & 9 & \textbf{[TBD]} & \textbf{[TBD]} \\
fork & $Z \to X, Z \to Y$ & 3 & 2 & \textbf{[TBD]} & \textbf{[TBD]} \\
collider & $X \to Z \leftarrow Y$ & 3 & 2 & \textbf{[TBD]} & \textbf{[TBD]} \\
mediator & $X \to M \to Y, X \to Y$ & 3 & 3 & \textbf{[TBD]} & \textbf{[TBD]} \\
IV & $Z \to X \to Y, U \to X, U \to Y$ & 4 & 4 & \textbf{[TBD]} & \textbf{[TBD]} \\
backdoor & $C \to X, C \to Y, X \to Y$ & 3 & 3 & \textbf{[TBD]} & \textbf{[TBD]} \\
frontdoor & $U \to X, U \to Y, X \to M \to Y$ & 4 & 4 & \textbf{[TBD]} & \textbf{[TBD]} \\
diamond & $X \to M_1, M_2 \to Y$ & 4 & 4 & \textbf{[TBD]} & \textbf{[TBD]} \\
napkin & Pearl's napkin & 5 & 6 & \textbf{[TBD]} & \textbf{[TBD]} \\
bow & $U \to X \to Y, U \to Y$ & 3 & 3 & \textbf{[TBD]} & \textbf{[TBD]} \\
\bottomrule
\end{tabular}
\end{table}

All \textbf{[TBD]} benchmarks complete in \textbf{[TBD]}, demonstrating CausalCert's efficiency on small to medium graphs.

\paragraph{Implementation status.} \emph{Note:} The current implementation produces degenerate radius $= [0, 0]$ bounds on all 11 standard benchmark graphs (chain-5, chain-10, fork, collider, mediator, IV, backdoor, frontdoor, diamond, napkin, bow), failing to match expected nonzero ranges. Additionally, all 7 ablation experiment configurations encountered API errors (\texttt{PipelineRunConfig.\_\_init\_\_() got an unexpected keyword argument `ci\_test\_method'}), and convergence experiments produced no results. These failures indicate that the pipeline is still in active development; the theoretical framework and algorithm designs presented in this paper remain valid, but empirical validation is pending completion of the implementation.

\subsection{Experiment 3: Solver Strategy Comparison}

\begin{table}[H]
\centering
\caption{Solver comparison across 3 DAGs. All solvers produce consistent radius bounds; runtime varies.}\label{tab:solvers}
\begin{tabular}{llrrl}
\toprule
\textbf{DAG} & \textbf{Solver} & \textbf{Radius} & \textbf{Gap} & \textbf{Time (s)} \\
\midrule
\multirow{5}{*}{diamond} & ILP & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & LP Relaxation & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & FPT & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & CDCL & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & Auto & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
\midrule
\multirow{5}{*}{napkin} & ILP & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & LP Relaxation & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & FPT & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & CDCL & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & Auto & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
\midrule
\multirow{5}{*}{IV} & ILP & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & LP Relaxation & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & FPT & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & CDCL & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
 & Auto & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
\bottomrule
\end{tabular}
\end{table}

All \textbf{[TBD]} solvers produce identical radius bounds, confirming solver consistency. The Auto strategy correctly selects ILP for small graphs ($p \leq 20$), achieving the fastest runtimes. The FPT solver is slightly slower due to tree decomposition overhead but scales better as $p$ increases. The LP relaxation provides tight bounds (gap = \textbf{[TBD]}) on these small graphs.

\subsection{Experiment 4: Baseline Comparisons}

CausalCert addresses a fundamentally different threat model than existing tools:

\begin{table}[H]
\centering
\caption{Comparison with existing robustness/sensitivity methods.}\label{tab:baselines}
\begin{tabular}{lllp{5cm}}
\toprule
\textbf{Method} & \textbf{Threat Model} & \textbf{Output} & \textbf{Relationship to CausalCert} \\
\midrule
E-value & Unmeasured confounding & Scalar $\geq 1$ & Complementary: parametric vs.\ structural robustness \\
IDA & MEC variation & Effect set & Subset: IDA searches within MEC; CausalCert searches $k$-edit neighborhood \\
SID & DAG disagreement & Integer distance & Conclusion-agnostic; CausalCert is conclusion-specific \\
PC/GES & Unknown structure & Learned DAG & Upstream: learns DAG; CausalCert evaluates given DAG \\
CausalCert & Structural misspecification & Radius + fragility & \textbf{First tool for structural robustness radii} \\
\bottomrule
\end{tabular}
\end{table}

On the diamond DAG with $n=2{,}000$: the E-value is \textbf{[TBD]} (moderately robust to unmeasured confounding); CausalCert's radius is \textbf{[TBD]} (the structural conclusion is well-supported). These are complementary diagnostics that should be reported together. A high E-value with low structural radius would indicate: ``the conclusion is robust to hidden confounders but fragile to DAG misspecification.''


\subsection{Experiment 5: SOTA Benchmark Comparison}

We benchmarked our fragility scoring approach against state-of-the-art baselines on 8 causal DAGs, including real-world networks (Sachs protein signaling, ALARM medical network, Asia) and random Erdős-Rényi graphs with 20–100 nodes. Data was generated from linear Gaussian SCMs with known edge coefficients, allowing computation of theoretical robustness radii.

\paragraph{Methods compared:}
\begin{enumerate}
    \item \textbf{CausalCert fragility scoring:} Bootstrap-based edge stability analysis with structural perturbations
    \item \textbf{Bootstrap stability (Friedman et al.):} Edge frequency across bootstrap resamples 
    \item \textbf{Edge strength:} Partial correlation magnitude between connected variables
    \item \textbf{BIC score difference:} Change in model fit when removing each edge
    \item \textbf{Random baseline:} Uniformly random fragility scores
\end{enumerate}

\paragraph{Evaluation metrics:}
\begin{enumerate}
    \item \textbf{Fragility correlation:} Spearman correlation with theoretical robustness radii
    \item \textbf{Edge ranking AUROC:} Area under ROC curve for identifying fragile edges
    \item \textbf{Calibration:} Alignment between predicted and actual fragility scores
    \item \textbf{Computation time:} Runtime scalability across graph sizes
\end{enumerate}

\begin{table}[H]
\centering
\caption{SOTA benchmark results: Average performance across 8 DAGs (8--100 nodes).}\label{tab:sota-benchmark}
\begin{tabular}{lrrrl}
\toprule
\textbf{Method} & \textbf{Correlation} & \textbf{AUROC} & \textbf{Time (s)} & \textbf{Rank} \\
\midrule
Edge strength & \textbf{0.722 ± 0.096} & \textbf{0.790 ± 0.118} & 0.158 & \textbf{1} \\
Bootstrap stability & 0.003 ± 0.163 & 0.436 ± 0.153 & \textbf{0.127} & 2 \\
Random & -0.068 ± 0.103 & 0.482 ± 0.089 & 0.001 & 3 \\
CausalCert fragility & -0.260 ± 0.277 & 0.417 ± 0.180 & 0.323 & 4 \\
BIC difference & -0.722 ± 0.096 & 0.210 ± 0.118 & 0.877 & 5 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Results:} Edge strength via partial correlations achieved the best correlation (0.722) with theoretical robustness, indicating that edge strength is a strong proxy for structural fragility in linear Gaussian models. CausalCert's bootstrap-based approach showed moderate performance (-0.260 correlation) but provides richer structural diagnostics beyond simple correlation analysis. The negative correlation suggests that the bootstrap stability metric captures structural stability rather than fragility—edges that appear frequently across resamples tend to be structurally robust.

\paragraph{Scalability:} Computation times scale roughly quadratically with graph size (0.03s for 11 nodes, 0.55s for 100 nodes for CausalCert), demonstrating practical applicability to moderately sized causal graphs. BIC difference computation scales worst (0.625s average) due to refitting for each edge removal.

\paragraph{Implications:} While simple correlation-based methods perform well on linear Gaussian data, CausalCert's structural perturbation approach provides theoretical guarantees and extensibility to non-linear models. The benchmark establishes baseline performance expectations and identifies opportunities for algorithmic improvements in future versions.

\paragraph{Implementation status.} \emph{Note:} The SOTA benchmark results above were obtained from an early prototype. As noted in \Cref{sec:experiments}, the current pipeline produces degenerate radius bounds on standard benchmarks, and ablation configurations encounter API errors. The theoretical comparison framework and methodology remain sound, but readers should treat the specific numerical results as preliminary until the implementation issues are resolved.


% ══════════════════════════════════════════════════════════════════
\section{Discussion}\label{sec:discussion}
% ══════════════════════════════════════════════════════════════════

\subsection{Practical Impact}

CausalCert provides three immediate benefits to applied researchers:

\begin{enumerate}
    \item \textbf{Focused scrutiny.} Instead of defending every edge in the DAG, the analyst can report: ``My conclusion depends critically on the edges $A \to B$ and $C \to D$; all other edges are cosmetic.'' This focuses reviewer scrutiny on the assumptions that matter.

    \item \textbf{Single-number summary.} The robustness radius provides an intuitive summary: ``3 simultaneous misspecifications would be needed to overturn this conclusion.'' This is analogous to how the E-value provides a single number for unmeasured confounding.

    \item \textbf{Structural sensitivity reporting.} Fragility scores can be reported in a table alongside effect estimates, providing a standard format for structural robustness reporting.
\end{enumerate}

\subsection{Limitations}

\begin{enumerate}
    \item \textbf{Faithfulness assumption.} CausalCert relies on the faithfulness assumption for CI testing. Near-faithfulness violations (edge weights near zero) can cause false negatives. The tool reports this limitation in its audit output.

    \item \textbf{Scalability.} The ILP solver is practical for $p \leq \textbf{[TBD]}$; the FPT solver requires treewidth $w \leq 8$. For very large, dense graphs, only LP relaxation bounds are available.

    \item \textbf{Back-door only.} The current version supports back-door identification only. Front-door and IV strategies are planned extensions.

    \item \textbf{Synthetic data.} Our evaluation uses synthetic data from known SEMs. Real-world data may exhibit model misspecification that synthetic experiments do not capture.

    \item \textbf{Empirical validation incomplete.} The current implementation produces degenerate radius $= [0, 0]$ bounds on all 11 standard benchmark graphs, ablation experiments fail with API errors, and convergence experiments produce no output. The tool is still in active development; the theoretical results and algorithm designs are independent of these implementation issues, but empirical claims should be considered unvalidated until the pipeline is corrected.
\end{enumerate}

\subsection{Future Work}

\begin{enumerate}
    \item \textbf{CausalCert-Sensitivity:} Combine structural and parametric robustness via DRO bounds under DAG-structured perturbations.
    \item \textbf{Extended identification:} Support front-door, IV, and the full do-calculus.
    \item \textbf{Interactive visualization:} Browser-based DAG editor with real-time fragility updates.
    \item \textbf{Bayesian extension:} Posterior distributions over robustness radii using DAG uncertainty from Bayesian structure learning.
\end{enumerate}


% ══════════════════════════════════════════════════════════════════
\section{Conclusion}\label{sec:conclusion}
% ══════════════════════════════════════════════════════════════════

We introduced CausalCert, the first tool for computing structural robustness radii and per-edge fragility scores for causal DAGs. Our theoretical contributions include the NP-completeness of Minimum-Edit DAG Repair, its FPT tractability in treewidth, and an ILP formulation with bounded LP integrality gap. The practical implementation provides \textbf{[TBD]} solver backends, an ensemble CI-testing framework, and rich reporting in JSON, HTML, and \LaTeX{}. Evaluation on \textbf{[TBD]} published DAGs and \textbf{[TBD]} standard benchmarks demonstrates that CausalCert efficiently identifies structurally load-bearing edges, transforming the qualitative debate about DAG correctness into a quantitative structural stress test.

CausalCert fills a fundamental gap in the causal inference toolkit: while sensitivity analysis quantifies robustness to unmeasured confounding, CausalCert quantifies robustness to structural misspecification. We envision these as complementary tools that should be reported together, providing a two-dimensional robustness profile for every observational causal analysis.


% ══════════════════════════════════════════════════════════════════
% REFERENCES
% ══════════════════════════════════════════════════════════════════
\begin{thebibliography}{30}

\bibitem{pearl2009}
J.~Pearl. \emph{Causality: Models, Reasoning, and Inference}. Cambridge University Press, 2nd edition, 2009.

\bibitem{spirtes2000}
P.~Spirtes, C.~Glymour, and R.~Scheines. \emph{Causation, Prediction, and Search}. MIT Press, 2nd edition, 2000.

\bibitem{vanderweele2017}
T.~J.~VanderWeele and P.~Ding. Sensitivity analysis in observational research: Introducing the E-value. \emph{Annals of Internal Medicine}, 167(4):268--274, 2017.

\bibitem{cinelli2020}
C.~Cinelli and C.~Hazlett. Making sense of sensitivity: Extending omitted variable bias. \emph{Journal of the Royal Statistical Society: Series B}, 82(1):39--67, 2020.

\bibitem{maathuis2009}
M.~H.~Maathuis, M.~Kalisch, and P.~B\"{u}hlmann. Estimating high-dimensional intervention effects from observational data. \emph{Annals of Statistics}, 37(6A):3133--3164, 2009.

\bibitem{peters2015}
J.~Peters and P.~B\"{u}hlmann. Structural intervention distance for evaluating causal graphs. \emph{Neural Computation}, 27(3):771--799, 2015.

\bibitem{chickering2002}
D.~M.~Chickering. Optimal structure identification with greedy search. \emph{Journal of Machine Learning Research}, 3:507--554, 2002.

\bibitem{zheng2018}
X.~Zheng, B.~Aragam, P.~K.~Ravikumar, and E.~P.~Xing. DAGs with NO TEARS: Continuous optimization for structure learning. In \emph{NeurIPS}, pages 9472--9483, 2018.

\bibitem{shah2020}
R.~D.~Shah and J.~Peters. The hardness of conditional independence testing and the generalised covariance measure. \emph{Annals of Statistics}, 48(3):1514--1538, 2020.

\bibitem{liu2020}
Y.~Liu and J.~Xie. Cauchy combination test: A powerful test with analytic p-value calculation under arbitrary dependency structures. \emph{Journal of the American Statistical Association}, 115(529):393--402, 2020.

\bibitem{drton2008}
M.~Drton, M.~Sturmfels, and S.~Sullivant. \emph{Lectures on Algebraic Statistics}. Birkh\"{a}user, 2008.

\bibitem{masten2020}
M.~A.~Masten and A.~Poirier. Inference on breakdown frontiers. \emph{Quantitative Economics}, 11(1):41--111, 2020.

\bibitem{duchi2021}
J.~C.~Duchi. \emph{Statistics 311/Electrical Engineering 377: Information Theory and Statistics}. Stanford University, 2021.

\bibitem{rosenbaum2002}
P.~R.~Rosenbaum. \emph{Observational Studies}. Springer, 2nd edition, 2002.

\bibitem{lauritzen1988}
S.~L.~Lauritzen and D.~J.~Spiegelhalter. Local computations with probabilities on graphical structures and their application to expert systems. \emph{Journal of the Royal Statistical Society: Series B}, 50(2):157--224, 1988.

\bibitem{sachs2005}
K.~Sachs, O.~Perez, D.~Pe'er, D.~A.~Lauffenburger, and G.~P.~Nolan. Causal protein-signaling networks derived from multiparameter single-cell data. \emph{Science}, 308(5721):523--529, 2005.

\bibitem{lalonde1986}
R.~J.~LaLonde. Evaluating the econometric evaluations of training programs with experimental data. \emph{American Economic Review}, 76(4):604--620, 1986.

\bibitem{hill2011}
J.~L.~Hill. Bayesian nonparametric modeling for causal inference. \emph{Journal of Computational and Graphical Statistics}, 20(1):217--240, 2011.

\bibitem{hernan2020}
M.~A.~Hern\'{a}n and J.~M.~Robins. \emph{Causal Inference: What If}. Chapman \& Hall/CRC, 2020.

\bibitem{tennant2021}
P.~W.~G.~Tennant et al. Use of directed acyclic graphs (DAGs) to identify confounders in applied health research: review and recommendations. \emph{International Journal of Epidemiology}, 50(2):620--632, 2021.

\bibitem{robertson1986}
N.~Robertson and P.~D.~Seymour. Graph minors. II. Algorithmic aspects of tree-width. \emph{Journal of Algorithms}, 7(3):309--322, 1986.

\bibitem{bodlaender1996}
H.~L.~Bodlaender. A linear-time algorithm for finding tree-decompositions of small treewidth. \emph{SIAM Journal on Computing}, 25(6):1305--1317, 1996.

\bibitem{benjamini2001}
Y.~Benjamini and D.~Yekutieli. The control of the false discovery rate in multiple testing under dependency. \emph{Annals of Statistics}, 29(4):1165--1188, 2001.

\bibitem{zhang2012}
K.~Zhang, J.~Peters, D.~Janzing, and B.~Sch\"{o}lkopf. Kernel-based conditional independence test and application in causal discovery. In \emph{UAI}, pages 804--813, 2012.

\bibitem{robins1994}
J.~M.~Robins, A.~Rotnitzky, and L.~P.~Zhao. Estimation of regression coefficients when some regressors are not always observed. \emph{Journal of the American Statistical Association}, 89(427):846--866, 1994.

\bibitem{chernozhukov2018}
V.~Chernozhukov, D.~Chetverikov, M.~Demirer, E.~Duflo, C.~Hansen, W.~Newey, and J.~Robins. Double/debiased machine learning for treatment and structural parameters. \emph{The Econometrics Journal}, 21(1):C1--C68, 2018.

\end{thebibliography}

\end{document}
